---
title: "Module 3 — Data visualization"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Mode flip.* If you were **agentic** in Exercise 1, you are **manual** now; if you were manual, you are agentic.

```{r setup}
library(tidyverse)
library(here)

table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Part 1 — Read two plots (no code)

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
ggplot(table_02, aes(x = ph, y = nugent_score, color = arm)) +
  geom_jitter()
```

**Plot 1 (above). Answer in plain words:**

1. What is the mark? Placebo and treatment arms are different colors, but they are the same mark.
2. What mark property comes from what column? The x and y positions of the points come from the `ph` and `nugent_score` columns, respectively. The color of the points comes from the `arm` column.
3. How many rows of data made each mark? Each point corresponds to one row of data, so the number of points equals the number of rows in the dataset.

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
ggplot(table_02, aes(x = arm, y = ph, fill = arm)) +
  geom_boxplot()
```

**Plot 2 (above). Answer the same three questions.**
1. What is the mark? Placebo and treatment arms are different colors, but they are the same mark.
2. What mark property comes from what column? The x and y positions of the points come from the `arm` and `ph` columns, respectively. The fill color of the boxes comes from the `arm` column.
3. How many rows of data made each mark? Each boxplot corresponds to one group (placebo or treatment), so the number of rows that contribute to each boxplot is the number of observations in each group. 

**Sketch.** Draw on paper or describe in words a plot that answers "does pH differ between treatment arms?" Name the mark, the mappings, and how many rows make each mark.

*Your sketch / description:*Mark: A boxplot for each treatment arm.
Mappings:
x position comes from arm (one box for placebo, one box for treatment).
y position comes from ph (vaginal pH values).
fill/color is also mapped to arm so the two boxes differ visually.
Rows per mark:
Each boxplot is drawn from all rows in that treatment group.
So one mark is made from all placebo rows, and the other mark from all treatment rows.

## Part 2 — Build the plot

Each step adds one piece on top of the previous plot. Run after each.

```{r}
# step 1: ggplot(table_02, aes(x = ph, y = nugent_score)) + geom_point()
ggplot(table_02, aes(x = ph, y = nugent_score)) +
  geom_point()
```

```{r}
# step 2: swap geom_point() for geom_jitter()
ggplot(table_02, aes(x = ph, y = nugent_score)) +
  geom_jitter()
```

```{r}
# step 3: add aes(color = arm)
ggplot(table_02, aes(x = ph, y = nugent_score, color = arm)) +
  geom_jitter()
```

```{r}
# step 4: switch to a boxplot of ph by arm: aes(x = arm, y = ph) + geom_boxplot()
ggplot(table_02, aes(x = arm, y = ph)) +
  geom_boxplot()
```

```{r}
# step 5a: aes(fill = arm) — mapping
ggplot(table_02, aes(x = arm, y = ph, fill = arm)) +
  geom_boxplot()
```

```{r}
# step 5b: fill = "navyblue" outside aes() — parameter
ggplot(table_02, aes(x = arm, y = ph)) +
  geom_boxplot(fill = "navyblue")
```

```{r}
# step 6: add labs(x = "Treatment arm", y = "Vaginal pH", fill = "Arm")
ggplot(table_02, aes(x = arm, y = ph, fill = arm)) +
  geom_boxplot() +
  labs(x = "Treatment arm", y = "Vaginal pH", fill = "Arm")
```

```{r}
# step 7: add facet_wrap(~ time_point)
ggplot(table_02, aes(x = arm, y = ph, fill = arm)) +
  geom_boxplot() +
  labs(x = "Treatment arm", y = "Vaginal pH", fill = "Arm") +
  facet_wrap(~ time_point)
```

## Part 3 — Reach (optional)

Produce a polished version of the final boxplot:

- faceted by `time_point`
- colored by `arm`
- axes and legend labeled
- panels in chronological order (baseline, week_1, week_7)
- a legible theme

```{r}
# reach: the polished final figure
ggplot(table_02, aes(x = arm, y = ph, fill = arm)) +
  geom_boxplot() +
  labs(x = "Treatment arm", y = "Vaginal pH", fill = "Arm") +
  facet_wrap(~ time_point, nrow = 1) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
