---
title: "Module 3 — Data wrangling"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi and let it write the code. Group B: **manual** — write each step yourself. A chat LLM may answer conceptual questions but should not edit this notebook or run commands. Exercise 2 flips the modes.

```{r setup}
library(tidyverse)
library(here)

table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Part 1 — Predict (no code)

Take a look at `table_02` (run the cell above first). Answer in plain words below — no code yet.

```{r}
table_02
```

**1. How many rows? What does one row represent?**

*Your answer:* 132 ROWS, each row represents a visit by a participant in the study.

**2. After `table_02 |> filter(time_point == "baseline")`, how many rows? What does each row represent now?**
```{r}
table_02|> filter(time_point == "baseline")
```
*Your answer:*44 rows, each row represents a baseline visit by a participant in the study.

**3. After `table_02 |> count(arm)`, how many rows? What does each row represent now?**

```{r}
table_02 |> count(arm)
```
*Your answer:*2 rows, each row represents a different arm of the study (intervention or control) and the count of participants in each arm.

## Part 2 — Build the pipeline

Each step adds one verb on top of the previous pipeline. Run after each step.

```{r}
head (table_02)  # step 1: pipe table_02 into head()
```

```{r}
# step 2: filter to the baseline time point
table_02 |> filter(time_point == "baseline")
```

```{r}
# step 3: add select() for pid, arm, ph, nugent_score, crp_blood
table_02 |> filter(time_point == "baseline") |> select(pid, arm, ph, nugent_score, crp_blood)
```

```{r}
# step 4: add mutate() for high_ph = if_else(ph >= 4.5, "yes", "no")
table_02 |> filter(time_point == "baseline") |> select(pid, arm, ph, nugent_score, crp_blood) |> mutate(high_ph = if_else(ph >= 4.5, "yes", "no"))
```

```{r}
# step 5: add arrange(desc(ph))
table_02 |> filter(time_point == "baseline") |> select(pid, arm, ph, nugent_score, crp_blood) |> mutate(high_ph = if_else(ph >= 4.5, "yes", "no")) |> arrange(desc(ph))
```

```{r}
# step 6: assign the final pipeline to early_visit, then print it
early_visit <- table_02 |> filter(time_point == "baseline") |> select(pid, arm, ph, nugent_score, crp_blood) |> mutate(high_ph = if_else(ph >= 4.5, "yes", "no")) |> arrange(desc(ph))
early_visit
```

```{r}
# step 7: separately, table_02 |> count(arm)
table_02 |> count(arm)
```

## Part 3 — Reach (optional)

Add a `nugent_category` column with `case_when()`:

- 0–3 → "normal"
- 4–6 → "intermediate"
- 7–10 → "BV"

Then run `count(arm, nugent_category)`.

```{r}
# reach: mutate(nugent_category = case_when(...)) then count(arm, nugent_category)
table_02 |> filter(time_point == "baseline") |> select(pid, arm, ph, nugent_score, crp_blood) |> mutate(high_ph = if_else(ph >= 4.5, "yes", "no"), nugent_category = case_when(nugent_score >= 0 & nugent_score <= 3 ~ "normal", nugent_score >= 4 & nugent_score <= 6 ~ "intermediate", nugent_score >= 7 & nugent_score <= 10 ~ "BV")) |> arrange(desc(ph)) |> count(arm, nugent_category)
```
