---
title: "Group A — Yogurt Study: Dataset Exploration"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

This document explores all four CSV files in the `data/group_a_yogurt/` folder.
The study looks at how eating yogurt affects the vaginal microbiome. Participants
were assigned to either a **yogurt** group or an **unchanged diet** group, and
samples were taken at **baseline** and **after antibiotic** treatment.

```{r setup, include=FALSE}
library(tidyverse)
```

---

## `00_sample_ids_yogurt.csv`

**What a row represents:** one biological sample collected from one participant
at one time point.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID (e.g. `pid_25`). Links to `01_participant_metadata_yogurt.csv` |
| `time_point` | When the sample was taken: `baseline` or `after_antibiotic` |
| `arm` | Study group assignment: `yogurt` or `unchanged_diet` |
| `sample_id` | Unique sample identifier (e.g. `UNCH001`). Links to `02_qpcr_results_yogurt.csv` and `03_luminex_results_yogurt.csv` |

```{r a00}
sample_ids <- read_csv("../../data/group_a_yogurt/00_sample_ids_yogurt.csv")
glimpse(sample_ids)
```

---

## `01_participant_metadata_yogurt.csv`

**What a row represents:** one study participant.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID. Links to `00_sample_ids_yogurt.csv` |
| `arm` | Study group: `yogurt` or `unchanged_diet` |
| `days_since_last_sex` | Number of days since the participant last had sex |
| `birth_control` | Type of birth control used: `no hormonal birth control` or `Depoprovera` |
| `age` | Age in years |
| `education` | Education level: `less than grade 9`, `grade 10-12, matriculated`, `grade 10-12, not matriculated`, or `post-secondary` |
| `sex` | Biological sex (all `female`) |

```{r a01}
participants <- read_csv("../../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
glimpse(participants)
```

---

## `02_qpcr_results_yogurt.csv`

**What a row represents:** one sample's bacterial DNA measurements from qPCR
(quantitative polymerase chain reaction — a lab technique that counts how many
bacterial DNA copies are present).

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_yogurt.csv` |
| `qpcr_bacteria` | Total bacterial load (number of DNA copies detected across all bacteria) |
| `qpcr_crispatus` | DNA copies of *Lactobacillus crispatus* — a "good" bacterium that keeps the vagina healthy |
| `qpcr_iners` | DNA copies of *Lactobacillus iners* — another *Lactobacillus* species, often associated with a less stable microbiome |

```{r a02}
qpcr <- read_csv("../../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
glimpse(qpcr)
```

---

## `03_luminex_results_yogurt.csv`

**What a row represents:** one cytokine measurement for one sample. Each sample
has multiple rows (one per cytokine tested). Cytokines are signalling proteins
the immune system uses to communicate — they tell us about inflammation levels.

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_yogurt.csv` |
| `cytokine` | Name of the cytokine measured: `IL-1a`, `IL-1b`, `IL-6`, `IL-8`, `IL-10`, `TNFa`, `IFN-Y`, `IP-10`, `MIG`, `MIP-3a` |
| `conc` | Concentration of the cytokine (in pg/mL — picograms per millilitre) |
| `limits` | Whether the value was `within limits` (reliable) or `out of range` (too low/high to measure accurately) |

```{r a03}
luminex <- read_csv("../../data/group_a_yogurt/03_luminex_results_yogurt.csv")
glimpse(luminex)
```

---

## Bringing it together

These four tables are linked by common ID columns:

```{r a04}
# How many participants in each group?
participants |>
  count(arm)

# How many samples per time point?
sample_ids |>
  count(time_point, arm)

# Are all samples matched with qPCR results?
sample_ids |>
  anti_join(qpcr, by = "sample_id")

# Join sample info to qPCR data for analysis
qpcr_joined <- sample_ids |>
  left_join(qpcr, by = "sample_id")

glimpse(qpcr_joined)

# Join sample info to Luminex data
luminex_joined <- sample_ids |>
  left_join(luminex, by = "sample_id")

glimpse(luminex_joined)
```