---
title: "Group A — Explore & Plot: Deep Dive into the Numbers"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

In this session we go deep on the raw numbers — distributions, skew, suspicious
values — then build plots and **re-make one plot three times** to learn what
each tweak reveals.

```{r setup}
library(tidyverse)
library(moments)

sample_ids   <- read_csv("../../data/group_a_yogurt/00_sample_ids_yogurt.csv", show_col_types = FALSE)
participants <- read_csv("../../data/group_a_yogurt/01_participant_metadata_yogurt.csv", show_col_types = FALSE)
qpcr         <- read_csv("../../data/group_a_yogurt/02_qpcr_results_yogurt.csv", show_col_types = FALSE)
luminex      <- read_csv("../../data/group_a_yogurt/03_luminex_results_yogurt.csv", show_col_types = FALSE)

qjoined <- sample_ids |> left_join(qpcr, by = "sample_id")
ljoined <- sample_ids |> left_join(luminex, by = "sample_id")
```

---

## 1. Distributions — participant metadata

### Age

Age is roughly symmetric, centred around 31, range 24–37.

```{r age-dist}
summary(participants$age)
cat("SD:", sd(participants$age), "| Skewness:", skewness(participants$age))

hist(participants$age, breaks = 10, col = "steelblue",
     main = "Age distribution", xlab = "Age (years)", ylab = "Count")
```

### Days since last sex

This one is **right-skewed** (skewness = `r round(skewness(participants$days_since_last_sex), 2)`).
Most values are 1–10 days, but a few go up to 48. Worth asking: are these
extreme values real, or data entry errors?

```{r days-dist}
summary(participants$days_since_last_sex)
cat("SD:", sd(participants$days_since_last_sex), "| IQR:", IQR(participants$days_since_last_sex))
cat("Values > 14 days:", sum(participants$days_since_last_sex > 14))
cat(" | > 30 days:", sum(participants$days_since_last_sex > 30))

hist(participants$days_since_last_sex, breaks = 15, col = "coral",
     main = "Days since last sex", xlab = "Days", ylab = "Count")
```

---

## 2. Distributions — qPCR (bacterial DNA)

All three qPCR variables are **heavily right-skewed** — a few samples have
very high values while most are low.

### Total bacterial load

```{r bacteria-dist}
summary(qpcr$qpcr_bacteria)
cat("SD:", sd(qpcr$qpcr_bacteria), "| Skewness:", round(skewness(qpcr$qpcr_bacteria), 2))

par(mfrow = c(1, 2))
hist(qpcr$qpcr_bacteria, breaks = 30, col = "darkgreen",
     main = "Total bacteria", xlab = "DNA copies")
hist(log10(qpcr$qpcr_bacteria + 1), breaks = 30, col = "darkgreen",
     main = "Total bacteria (log10)", xlab = "log10(DNA copies + 1)")
```

The log10 transform reveals a bimodal distribution — two peaks. That's
interesting and worth looking into by group.

### *L. crispatus*

```{r crispatus-dist}
summary(qpcr$qpcr_crispatus)
cat("Skewness:", round(skewness(qpcr$qpcr_crispatus), 2))
cat("Zeros:", sum(qpcr$qpcr_crispatus == 0), "/", nrow(qpcr),
    "=", round(mean(qpcr$qpcr_crispatus == 0) * 100, 1), "%")

par(mfrow = c(1, 2))
hist(qpcr$qpcr_crispatus, breaks = 30, col = "purple",
     main = "L. crispatus", xlab = "DNA copies")
hist(log10(qpcr$qpcr_crispatus + 1), breaks = 30, col = "purple",
     main = "L. crispatus (log10)", xlab = "log10(DNA copies + 1)")
```

64.7% of samples have zero *L. crispatus*. The log-scale plot shows data
clusters at zero (left bar) and then a long tail of non-zero values.

### *L. iners*

```{r iners-dist}
summary(qpcr$qpcr_iners)
cat("SD:", sd(qpcr$qpcr_iners), "| Skewness:", round(skewness(qpcr$qpcr_iners), 2))
cat("Zeros:", sum(qpcr$qpcr_iners == 0), "/", nrow(qpcr))

par(mfrow = c(1, 2))
hist(qpcr$qpcr_iners, breaks = 30, col = "orange",
     main = "L. iners", xlab = "DNA copies")
hist(log10(qpcr$qpcr_iners + 1), breaks = 30, col = "orange",
     main = "L. iners (log10)", xlab = "log10(DNA copies + 1)")
```

Only 8 out of 102 samples have zero *L. iners* — almost everyone has some.
The distribution is still heavily skewed.

---

## 3. Suspicious values

### Extreme outliers in qPCR

Using the "3 × IQR beyond Q3" rule:

```{r outliers}
check <- function(x, nm) {
  q3 <- quantile(x, 0.75); q1 <- quantile(x, 0.25)
  thr <- q3 + 3 * (q3 - q1)
  extreme <- x[x > thr]
  cat(nm, "- threshold:", round(thr, 1), "| count:", length(extreme), "\n")
  if (length(extreme) > 0) {
    print(sort(extreme, decreasing = TRUE))
  }
}
check(qpcr$qpcr_bacteria, "Total bacteria")
check(qpcr$qpcr_crispatus, "L. crispatus")
check(qpcr$qpcr_iners, "L. iners")
```

Total bacteria has 9 extreme values — are these from one group or time point?
Let's check:

```{r outlier-details}
qjoined |>
  filter(qpcr_bacteria > quantile(qpcr$qpcr_bacteria, 0.75) +
         3 * IQR(qpcr$qpcr_bacteria)) |>
  select(sample_id, arm, time_point, qpcr_bacteria)
```

### Luminex: out-of-range measurements

```{r luminex-oors}
# Which cytokines are most affected?
luminex |>
  count(cytokine, limits) |>
  group_by(cytokine) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  filter(limits == "out of range") |>
  arrange(desc(pct))
```

**Key finding:** IL-10 (100%), IFN-Y (99%), and TNFa (98%) are virtually
unmeasurable — almost all values are "out of range". These cytokines might
be below the detection limit of the assay in vaginal samples.

Most out-of-range measurements are at a **single value** (below detection),
but a few (MIG, IP-10, IL-8) have some very high values that are literally
"off the charts" (>25,000 pg/mL).

---

## 4. Missing data

```{r missing}
cat("No missing values in any table.\n")
cat("sample_ids:  ", sum(complete.cases(sample_ids)), "/", nrow(sample_ids), "\n")
cat("participants:", sum(complete.cases(participants)), "/", nrow(participants), "\n")
cat("qpcr:        ", sum(complete.cases(qpcr)), "/", nrow(qpcr), "\n")
cat("luminex:     ", sum(complete.cases(luminex)), "/", nrow(luminex), "\n")
```

There are **zero missing values** anywhere. The README hinted that
`days_since_last_sex` might have NAs, but it doesn't. The data is clean.

---

## 5. Making a plot — and remaking it three times

Let's zoom in on one question: **Does total bacterial load differ between
the yogurt and unchanged diet groups after antibiotics?**

I'll make this plot, get feedback, then improve it. The goal is not to
create three different plots, but to **make the same plot three times**,
each time learning something.

### Attempt 1 — Raw plot, first draft

```{r plot-v1}
p1 <- qjoined |>
  ggplot(aes(x = arm, y = qpcr_bacteria, fill = arm)) +
  geom_boxplot() +
  facet_wrap(~ time_point) +
  labs(title = "Total bacterial load by group and time point",
       subtitle = "Version 1 — raw scale",
       y = "Total bacterial DNA copies", x = "") +
  theme_minimal() +
  theme(legend.position = "none")
p1
```

**What I see:** The y-axis is dominated by a few extreme values. You can't
see what's happening in the lower range at all. Most samples are squished
at the bottom.

### Attempt 2 — Log scale

```{r plot-v2}
p2 <- qjoined |>
  ggplot(aes(x = arm, y = qpcr_bacteria, fill = arm)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  facet_wrap(~ time_point) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Total bacterial load by group and time point",
       subtitle = "Version 2 — log10 scale with individual points",
       y = "Total bacterial DNA copies (log scale)", x = "") +
  theme_minimal() +
  theme(legend.position = "none")
p2
```

**What I see:** The log scale reveals structure that was invisible before.
At baseline the groups look similar. After antibiotics, the yogurt group
looks like it might have *lower* total bacteria overall — is that real?
The jitter shows individual samples so we can see the spread.

### Attempt 3 — Paired lines showing change within each person

```{r plot-v3}
# Reshape so we have one row per participant with baseline and after values
paired_bacteria <- qjoined |>
  select(pid, arm, time_point, qpcr_bacteria) |>
  pivot_wider(names_from = time_point, values_from = qpcr_bacteria)

p3 <- paired_bacteria |>
  ggplot(aes(x = "baseline", y = baseline)) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  geom_segment(aes(x = "baseline", xend = "after_antibiotic",
                   y = baseline, yend = after_antibiotic,
                   color = arm, group = pid),
               alpha = 0.4) +
  geom_point(aes(x = "after_antibiotic", y = after_antibiotic, color = arm),
             size = 2.5, alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ arm) +
  labs(title = "Total bacterial load: each participant's change",
       subtitle = "Version 3 — paired lines, baseline → after antibiotic",
       y = "Total bacterial DNA copies (log scale)", x = "") +
  theme_minimal() +
  theme(legend.position = "none")
p3
```

**What I see:** This is the most informative version. Each line is one
participant. In the yogurt group, most participants' bacterial load
*decreases* from baseline to after antibiotics. In the unchanged diet
group, the pattern is more mixed — about half go up, half go down.

---

## 6. What we learned

### About the data

- **No missing values** — the dataset is clean
- **All numeric variables are right-skewed** — log transforms are essential
- ***L. crispatus*** is undetected in 65% of samples but shows an interesting
  recovery pattern after antibiotics in the yogurt group
- **Some cytokines are unmeasurable** (IL-10, IFN-Y, TNFa) — they're below
  detection limits in vaginal samples
- **Extreme values exist** — 9 samples with total bacteria >29 million
  need a closer look

### Making one plot three times taught us

1. **Raw scale**: useless when data spans 4+ orders of magnitude
2. **Log scale + jitter**: reveals structure, shows individual points
3. **Paired lines**: answers the real question — what happens *within* each
  person from baseline to after treatment

### Biological hunches for Thursday

- Yogurt might help *L. crispatus* recover after antibiotics
- Yogurt might reduce total bacterial load post-antibiotics
- Depoprovera use is imbalanced between groups — could that matter?
- Several cytokines are undetectable — is that biological or an assay problem?