---
title: "Group A — Yogurt Study: Final Presentation"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(gtsummary)
library(ggpubr)

sample_ids   <- read_csv("../../data/group_a_yogurt/00_sample_ids_yogurt.csv")
participants <- read_csv("../../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
qpcr         <- read_csv("../../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
luminex      <- read_csv("../../data/group_a_yogurt/03_luminex_results_yogurt.csv")

qpcr_joined    <- sample_ids |> left_join(qpcr, by = "sample_id")
luminex_joined <- sample_ids |> left_join(luminex, by = "sample_id")
```

---

## 1. What biological question did you investigate?

> **Does eating yogurt help *Lactobacillus crispatus* — the protective
> vaginal bacterium — recover after antibiotic treatment?**
>
> Antibiotics disrupt the vaginal microbiome by killing bacteria
> indiscriminately. We wanted to know if eating yogurt (which contains
> bacteria) helps *L. crispatus* bounce back faster compared to an
> unchanged diet.

---

## 2. Who are the participants?

```{r table-one}
participants |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label = "**Characteristic**")
```

**Key takeaways:**

- **51 participants** (25 unchanged diet, 26 yogurt) — roughly balanced
- **Age similar** across groups (mean ≈ 31 in both, p = 0.55)
- But **birth control is imbalanced**: 17 of 26 yogurt participants use
  Depoprovera vs. only 8 of 25 in the unchanged diet group. This could
  matter for the microbiome.

---

## 3. What did you compare, and what did you find?

### Main finding: *L. crispatus* recovery after antibiotics

In the **yogurt group**, 85% of samples had detectable *L. crispatus* after
antibiotics. In the **unchanged diet group**, only 48% did.

```{r crispatus-recovery}
# Prepare paired data for L. crispatus
qpcr_joined |>
  filter(time_point %in% c("baseline", "after_antibiotic")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic")) |>
  ggplot(aes(time_point, qpcr_crispatus)) +
  geom_line(aes(group = pid), alpha = 0.3) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  facet_wrap(~ arm) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "L. crispatus: baseline → after antibiotic (each line = one person)",
    subtitle = "Yogurt group shows more recovery after antibiotics",
    y = "L. crispatus DNA copies (log scale)",
    x = "",
    color = "Group"
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = list(c("baseline", "after_antibiotic"))
  ) +
  theme_minimal()
```

**What this shows:** Each line is one participant. At baseline, almost
everyone has near-zero *L. crispatus* in both groups. After antibiotics,
the yogurt group has many participants with visible *L. crispatus* levels,
while the unchanged diet group mostly stays flat at zero.

### Secondary finding: Total bacterial load

```{r total-bacteria}
paired_bacteria <- qpcr_joined |>
  select(pid, arm, time_point, qpcr_bacteria) |>
  pivot_wider(names_from = time_point, values_from = qpcr_bacteria)

paired_bacteria |>
  ggplot(aes(x = "Baseline", y = baseline)) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  geom_segment(aes(x = "Baseline", xend = "After antibiotic",
                   y = baseline, yend = after_antibiotic,
                   color = arm, group = pid),
               alpha = 0.4) +
  geom_point(aes(x = "After antibiotic", y = after_antibiotic, color = arm),
             size = 2.5, alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ arm) +
  labs(
    title = "Total bacterial load: each participant's change",
    subtitle = "Yogurt group tends to decrease; unchanged diet is mixed",
    y = "Total bacterial DNA copies (log scale)",
    x = "",
    color = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**What this shows:** Most participants in the yogurt group show a
*decrease* in total bacterial load from baseline to after antibiotics.
The unchanged diet group is more mixed — about half go up, half go down.

---

## 4. Limitations

> **What could go wrong with our conclusion?**
>
> 1. **Birth control imbalance** — 65% of the yogurt group uses Depoprovera
>    vs. 32% of the unchanged diet group. Depoprovera affects hormones and
>    could independently influence the vaginal microbiome. The *L. crispatus*
>    "recovery" might partly reflect this difference, not the yogurt.
>
> 2. **No placebo** — participants knew which group they were in. We can't
>    rule out placebo effects or behavioural changes.
>
> 3. ***L. crispatus* mostly absent at baseline** — with 96% of samples
>    showing zero at baseline, we're comparing recovery from near-zero,
>    which is hard to measure precisely.
>
> 4. **Small sample** — 51 participants is modest. One or two extreme
>    values can influence the results.
>
> 5. **Out-of-range cytokine measurements** — several cytokines (IL-10,
>    IFN-Y, TNFa) were almost entirely below detection limits, so we
>    couldn't assess immune response for those.

---

## 5. If you had another week, what would you do next?

> 1. **Adjust for Depoprovera** — run a model that controls for birth
>    control type to see if the yogurt effect holds up
> 2. **Look at *L. iners*** — does yogurt push the microbiome toward
>    *L. crispatus* dominance and away from *L. iners*?
> 3. **Check the usable cytokines** (IL-1a, IL-8, IP-10, MIG, MIP-3a)
>    — is there a difference in inflammation between groups?
> 4. **Identify the 9 extreme total bacteria samples** — are they from
>    the same participants? Do they cluster in one group?
> 5. **Build a simple model** predicting *L. crispatus* recovery from
>    group + age + birth control

---

## 6. What did your group learn this week?

> - **Data mapping first**: Understanding what one row means in each table
>   and how tables connect saved us from making mistakes later
> - **Log scales are not optional**: When data spans 4+ orders of magnitude
>   (like bacterial DNA counts), raw-scale plots hide everything
> - **Zeros can be meaningful**: 65% of samples had zero *L. crispatus* —
>   that's not missing data, that's biology
> - **"Out of range" ≠ unusable**: Many cytokine measurements were below
>   detection limits, which tells us the assay may not be sensitive enough
>   for vaginal samples
> - **Making one plot three times** taught us more than making three
>   different plots once — each iteration revealed something new

---

## Bonus: Shortlist of biological questions for Thursday

1. **Does yogurt restore *L. crispatus* after antibiotics, controlling
   for Depoprovera?** (Our main finding, but needs adjustment)
2. **Does the yogurt group have a different *L. crispatus* / *L. iners*
   ratio after antibiotics?** (A more stable microbiome has more *L. crispatus*)
3. **Is total bacterial load reduction after antibiotics beneficial or
   harmful?** (More bacteria isn't always better)
4. **Do baseline characteristics (age, birth control, days since last sex)
   predict microbiome recovery regardless of group?**
5. **Are the extreme total bacteria values real biological variation or
   sample contamination?** (Samples >29 million copies need investigation)
