---
title: "Group A — Yogurt & the Vaginal Microbiome"
subtitle: "Does eating yogurt help *L. crispatus* recover after antibiotics?"
author: "Group A"
date: "Friday, 5 June 2026"
format:
  revealjs:
    theme: default
    slide-number: true
    code-fold: false
    execute:
      warning: false
      message: false
    embed-resources: true
    toc: false
    self-contained: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(gtsummary)
library(ggpubr)

sample_ids   <- read_csv("../../data/group_a_yogurt/00_sample_ids_yogurt.csv")
participants <- read_csv("../../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
qpcr         <- read_csv("../../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
luminex      <- read_csv("../../data/group_a_yogurt/03_luminex_results_yogurt.csv")

qpcr_joined <- sample_ids |> left_join(qpcr, by = "sample_id")
```

## Our question

> **Does eating yogurt help *Lactobacillus crispatus* — the "good" vaginal bacterium — recover after antibiotic treatment?**

- Antibiotics kill bacteria indiscriminately, disrupting the vaginal microbiome
- Yogurt contains live bacteria — could it help *L. crispatus* bounce back?
- **51 participants** assigned to either **yogurt** or **unchanged diet**
- Sampled at **baseline** (before study) and **after antibiotic** treatment

---

## Who are the participants?

```{r table1}
participants |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test")) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label = "**Characteristic**")
```

### Key takeaways
- **51 women** — 25 unchanged diet, 26 yogurt
- Age is balanced (≈31 in both groups, p = 0.55)
- ⚠️ **Birth control is imbalanced** — 65% of yogurt group uses Depoprovera vs. 32% of unchanged diet

---

## What did we find? *L. crispatus* recovery

```{r main-figure}
qpcr_joined |>
  filter(time_point %in% c("baseline", "after_antibiotic")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic")) |>
  ggplot(aes(time_point, qpcr_crispatus)) +
  geom_line(aes(group = pid), alpha = 0.3) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  facet_wrap(~ arm) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "L. crispatus: baseline → after antibiotic",
    subtitle = "Each line = one participant",
    y = "L. crispatus DNA copies (log scale)",
    x = "",
    color = "Group"
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = list(c("baseline", "after_antibiotic"))
  ) +
  theme_minimal()
```

---

## What the figure shows

`r qpcr_joined |> filter(time_point == "after_antibiotic") |> group_by(arm) |> summarise(n_zero = sum(qpcr_crispatus == 0), n = n(), pct_detected = round((1 - n_zero/n)*100)) |> mutate(group = arm) |> select(group, pct_detected) |> knitr::kable(col.names = c("Group", "% with detectable L. crispatus"))`

| Time point | Yogurt group | Unchanged diet |
|-----------|-------------|----------------|
| **Baseline** | 96% undetected | 96% undetected |
| **After antibiotic** | **85% detected** | **48% detected** |

- At baseline: almost everyone has zero *L. crispatus* in both groups
- After antibiotics: **the yogurt group shows clear recovery**; the unchanged diet group mostly stays at zero
- Paired Wilcoxon tests confirm within-group change is significant for yogurt (p < 0.001) but not for unchanged diet

---

## Limitations

> **What could go wrong with our conclusion?**

1. **Birth control imbalance** — 17/26 yogurt participants use Depoprovera vs. 8/25 in unchanged diet. Depoprovera affects hormones and could independently influence the microbiome
2. **No placebo** — participants knew their group
3. **Small sample** — 51 participants; a few extreme values can shift results
4. **No adjustment** — we haven't controlled for confounders yet
5. **Cytokines mostly unmeasurable** — several inflammatory markers were below detection limits, so we couldn't check immune response

---

## If we had another week

1. **Adjust for Depoprovera** in the model — does the yogurt effect hold?
2. **Check *L. iners*** — does yogurt shift the *L. crispatus* / *L. iners* ratio?
3. **Identify the 9 extreme samples** with very high bacterial load — are they from one group?
4. **Build a simple linear model** predicting *L. crispatus* from group + age + birth control

---

## What we learned

> **From the data**
> - Log scales aren't optional when data spans 4+ orders of magnitude
> - Zeros aren't missing — 65% of samples had zero *L. crispatus*, which is real biology
> - "Out of range" cytokine measurements (IL-10, IFN-Y, TNFa) tell us the assay may not be sensitive enough for vaginal samples

> **From the process**
> - Making one plot three times teaches more than making three different plots once
> - Always verify: does the test match the distribution? (Skewed data → Wilcoxon)
> - A clean dataset with no missing values still needs careful interpretation

---

## Thank you

**Any questions?**

Slides built with Quarto + revealjs

Code and data: `data/group_a_yogurt/` and `group_project/group_a/`