---
title: "Group A — Yogurt Study: Table 1 and Group Comparisons"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

In this session we'll pick a biological question, build a **Table 1** to
describe the participants (yogurt vs. unchanged diet), then compare groups
using **gtsummary** and **ggpubr** with p-values.

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)
library(ggpubr)

# Load all four datasets
sample_ids  <- read_csv("../../data/group_a_yogurt/00_sample_ids_yogurt.csv")
participants <- read_csv("../../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
qpcr        <- read_csv("../../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
luminex     <- read_csv("../../data/group_a_yogurt/03_luminex_results_yogurt.csv")

# Join sample info to qPCR and Luminex
qpcr_joined <- sample_ids |>
  left_join(qpcr, by = "sample_id")

luminex_joined <- sample_ids |>
  left_join(luminex, by = "sample_id")
```

---

## Step 1 — What biological question interests your group?

Write your group's question here:

> *Your question:* ___________________________________________________________________

---

## Step 2 — Build a Table 1

A Table 1 summarises the demographic and baseline characteristics of
participants, split by study group. This lets us check if the groups are
similar at the start of the study.

```{r step2}
# tbl_summary with defaults
participants |>
  select(-pid, -sex) |>  # remove ID and constant columns
  tbl_summary(by = arm)
```

```{r step2b}
# Add p-values to see if groups already differ at baseline
participants |>
  select(-pid, -sex) |>
  tbl_summary(by = arm) |>
  add_p()
```

```{r step2c}
# Polished version with overall column and bold labels
participants |>
  select(-pid, -sex) |>
  tbl_summary(by = arm) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label = "**Characteristic**")
```

```{r step2d}
# Override: show age as mean (SD) with a t-test
participants |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test")) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label = "**Characteristic**")
```

**Interpretation:** Are the two groups similar at baseline? Any variables where
p < 0.05? What does that mean for our study?

---

## Step 3 — Compare outcomes between groups

Now let's look at the actual study outcomes. We'll compare the yogurt group
vs. the unchanged diet group using different bacterial and cytokine measures.

### 3a — Total bacterial load (qPCR)

```{r step3a}
# Boxplot of total bacterial load at each time point
qpcr_joined |>
  ggplot(aes(arm, qpcr_bacteria, fill = arm)) +
  geom_boxplot() +
  facet_wrap(~ time_point) +
  labs(
    title = "Total bacterial load by group",
    subtitle = "Faceted by time point (baseline vs. after antibiotic)",
    y = "Total bacterial DNA copies",
    x = ""
  ) +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

### 3b — *L. crispatus* (protective bacterium)

```{r step3b}
qpcr_joined |>
  ggplot(aes(arm, qpcr_crispatus, fill = arm)) +
  geom_boxplot() +
  facet_wrap(~ time_point) +
  scale_y_log10() +  # log scale because data spans many orders of magnitude
  labs(
    title = "Lactobacillus crispatus by group",
    subtitle = "Y-axis on log scale",
    y = "L. crispatus DNA copies (log scale)",
    x = ""
  ) +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

### 3c — *L. iners* (less stable species)

```{r step3c}
qpcr_joined |>
  ggplot(aes(arm, qpcr_iners, fill = arm)) +
  geom_boxplot() +
  facet_wrap(~ time_point) +
  scale_y_log10() +
  labs(
    title = "Lactobacillus iners by group",
    y = "L. iners DNA copies (log scale)",
    x = ""
  ) +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

### 3d — Cytokine levels (pick a few)

```{r step3d}
# Panel of key cytokines, only within-limits measurements
cytokines_of_interest <- c("IL-1a", "IL-1b", "IL-6", "IL-8", "IL-10", "TNFa")

luminex_joined |>
  filter(cytokine %in% cytokines_of_interest, limits == "within limits") |>
  ggplot(aes(arm, conc, fill = arm)) +
  geom_boxplot() +
  facet_wrap(~ cytokine, scales = "free_y") +
  labs(
    title = "Cytokine concentrations by group",
    subtitle = "Only within-limits measurements shown",
    y = "Concentration (pg/mL)",
    x = ""
  ) +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

---

## Step 4 — Paired analysis: baseline vs. after antibiotic

Each participant was sampled at two time points, so we can also look at how
things *change within* each group.

```{r step4}
# Prepare the data for paired analysis
qpcr_paired <- qpcr_joined |>
  filter(time_point %in% c("baseline", "after_antibiotic")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic"))

# Line plot: total bacteria, pre vs. post, by group
qpcr_paired |>
  ggplot(aes(time_point, qpcr_bacteria)) +
  geom_line(aes(group = pid), alpha = 0.3) +
  geom_point(aes(color = arm), size = 2) +
  facet_wrap(~ arm) +
  labs(
    title = "Change in total bacterial load: baseline → after antibiotic",
    y = "Total bacterial DNA copies",
    x = ""
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = list(c("baseline", "after_antibiotic"))
  )
```

---

## Step 5 (Optional) — Adjusted analysis

We can also run a simple linear model that adjusts for baseline differences.

```{r step5, eval=FALSE}
# Only run this if you want to adjust for age or other variables
outcome <- qpcr_joined |>
  filter(time_point == "after_antibiotic") |>
  left_join(participants, by = c("pid", "arm"))

lm(log10(qpcr_bacteria) ~ arm + age, data = outcome) |>
  tbl_regression()
```

---

## What does your group conclude?

Write your takeaways here:

> *Our findings:*
> _______________________________________________________________________
> _______________________________________________________________________
> _______________________________________________________________________