this ---
title: "Group 1 — Dataset Exploration"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
```

This document explores all the CSV files found in the `data/` folder. There are
four distinct datasets, each in its own subfolder. For every file we'll describe
what a single row represents and what each column means.

---

# Dataset A — Yogurt Study (`data/group_a_yogurt/`)

A study looking at how eating yogurt affects the vaginal microbiome. Participants
were assigned to either a **yogurt** group or an **unchanged diet** group, and
samples were taken at **baseline** and **after antibiotic** treatment.

## `00_sample_ids_yogurt.csv`

**What a row represents:** one biological sample collected from one participant
at one time point.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID (e.g. `pid_25`). Links to `01_participant_metadata_yogurt.csv` |
| `time_point` | When the sample was taken: `baseline` or `after_antibiotic` |
| `arm` | Study group assignment: `yogurt` or `unchanged_diet` |
| `sample_id` | Unique sample identifier (e.g. `UNCH001`). Links to `02_qpcr_results_yogurt.csv` and `03_luminex_results_yogurt.csv` |

```{r a00}
sample_ids <- read_csv("../../data/group_a_yogurt/00_sample_ids_yogurt.csv")
glimpse(sample_ids)
```

---

## `01_participant_metadata_yogurt.csv`

**What a row represents:** one study participant.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID. Links to `00_sample_ids_yogurt.csv` |
| `arm` | Study group: `yogurt` or `unchanged_diet` |
| `days_since_last_sex` | Number of days since the participant last had sex |
| `birth_control` | Type of birth control used: `no hormonal birth control` or `Depoprovera` |
| `age` | Age in years |
| `education` | Education level: `less than grade 9`, `grade 10-12, matriculated`, `grade 10-12, not matriculated`, or `post-secondary` |
| `sex` | Biological sex (all `female`) |

```{r a01}
participants_yogurt <- read_csv("../../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
glimpse(participants_yogurt)
```

---

## `02_qpcr_results_yogurt.csv`

**What a row represents:** one sample's bacterial DNA measurements from qPCR
(quantitative polymerase chain reaction — a lab technique that counts how many
bacterial DNA copies are present).

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_yogurt.csv` |
| `qpcr_bacteria` | Total bacterial load (number of DNA copies detected across all bacteria) |
| `qpcr_crispatus` | DNA copies of *Lactobacillus crispatus* — a "good" bacterium that keeps the vagina healthy |
| `qpcr_iners` | DNA copies of *Lactobacillus iners* — another *Lactobacillus* species, often associated with a less stable microbiome |

```{r a02}
qpcr <- read_csv("../../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
glimpse(qpcr)
```

---

## `03_luminex_results_yogurt.csv`

**What a row represents:** one cytokine measurement for one sample. Each sample
has multiple rows (one per cytokine tested). Cytokines are signalling proteins
the immune system uses to communicate — they tell us about inflammation levels.

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_yogurt.csv` |
| `cytokine` | Name of the cytokine measured: `IL-1a`, `IL-1b`, `IL-6`, `IL-8`, `IL-10`, `TNFa`, `IFN-Y`, `IP-10`, `MIG`, `MIP-3a` |
| `conc` | Concentration of the cytokine (in pg/mL — picograms per millilitre) |
| `limits` | Whether the value was `within limits` (reliable) or `out of range` (too low/high to measure accurately) |

```{r a03}
luminex_yogurt <- read_csv("../../data/group_a_yogurt/03_luminex_results_yogurt.csv")
glimpse(luminex_yogurt)
```

---

# Dataset B — Menstruation Study (`data/group_b_menstruation/`)

A study of the vaginal microbiome throughout the menstrual cycle, comparing
participants who use **birth control** vs **no birth control**. Samples were
taken at four time points across the cycle.

## `00_sample_ids_period.csv`

**What a row represents:** one biological sample from one participant at one
menstrual-cycle time point.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID. Links to `01_participant_metadata_period.csv` |
| `time_point` | When in the cycle: `onset` (period starts), `end_bleeding`, `week_post`, or `week_prior` |
| `arm` | Study group: `birth_control` or `no_birth_control` |
| `sample_id` | Unique sample identifier. Links to `02_luminex_period.csv` and `03_flow_cytometry_period.csv` |

```{r b00}
sample_ids_period <- read_csv("../../data/group_b_menstruation/00_sample_ids_period.csv")
glimpse(sample_ids_period)
```

---

## `01_participant_metadata_period.csv`

**What a row represents:** one study participant.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID |
| `arm` | Study group: `birth_control` or `no_birth_control` |
| `age` | Age in years |
| `pcos_status` | Whether the participant has PCOS (polycystic ovary syndrome): `pcos` or `no disease` |
| `period_product` | Menstrual product used: `tampon`, `pad`, or `menstrual_cup` |
| `sex` | Biological sex (all `female`) |

```{r b01}
participants_period <- read_csv("../../data/group_b_menstruation/01_participant_metadata_period.csv")
glimpse(participants_period)
```

---

## `02_luminex_period.csv`

**What a row represents:** one cytokine measurement for one sample (long format —
each sample has multiple rows, one per cytokine tested).

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_period.csv` |
| `cytokine` | Cytokine name: `IFNg`, `IL-1a`, `IL-1b`, `IL-6`, `IP-10`, `MIG`, `MIP-3a`, `TNFa` |
| `conc` | Concentration in pg/mL |
| `limits` | Measurement quality: `not_censored` (reliable), `below_detection_limit`, or `above_detect_limit` |

```{r b02}
luminex_period <- read_csv("../../data/group_b_menstruation/02_luminex_period.csv")
glimpse(luminex_period)
```

---

## `03_flow_cytometry_period.csv`

**What a row represents:** one sample's immune cell counts from flow cytometry
(a lab technique that counts and classifies cells using lasers).

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_period.csv` |
| `live_cd19_negative` | Number of live cells that are not B-cells (CD19 is a B-cell marker) |
| `cd45_negative` | Cells that do not express CD45 — mostly non-immune cells |
| `cd45_positive` | Cells that express CD45 — these are immune cells (white blood cells) |
| `neutrophils` | Number of neutrophils — a type of immune cell that fights bacteria |
| `non_neutrophils` | Immune cells that are not neutrophils |
| `cd3_negative` | Immune cells that are not T-cells (CD3 is a T-cell marker) |
| `cd3_positive` | T-cells (a key type of lymphocyte) |
| `cd4_t_cells` | Helper T-cells — they coordinate immune responses |
| `cd8_t_cells` | Killer T-cells — they destroy infected cells |

```{r b03}
flow <- read_csv("../../data/group_b_menstruation/03_flow_cytometry_period.csv")
glimpse(flow)
```

---

# Dataset C — Instructional Dataset (`data/instructional_dataset/`)

A teaching dataset from the UKZN workshop 2023 — a study where participants
received either a **treatment** or a **placebo**, with samples collected at
**baseline**, **week 1**, and **week 7**.

## `00_sample_ids_UKZN_workshop_2023.csv`

**What a row represents:** one biological sample from one participant at one
study visit.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID. Links to `01_participant_metadata_UKZN_workshop_2023.csv` and `02_visit_clinical_measurements_UKZN_workshop_2023.csv` |
| `time_point` | Study visit: `baseline`, `week_1`, or `week_7` |
| `arm` | Study group: `treatment` or `placebo` |
| `sample_id` | Unique sample identifier. Links to `03_elisa_cytokines_UKZN_workshop_2023.csv` and `04_flow_cytometry_UKZN_workshop_2023.csv` |

```{r c00}
sample_ids_inst <- read_csv("../../data/instructional_dataset/00_sample_ids_UKZN_workshop_2023.csv")
glimpse(sample_ids_inst)
```

---

## `01_participant_metadata_UKZN_workshop_2023.csv`

**What a row represents:** one study participant.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID |
| `arm` | Study group: `treatment` or `placebo` |
| `smoker` | Smoking status: `smoker` or `non-smoker` |
| `age` | Age in years |
| `education` | Education level: `less than grade 9`, `grade 10-12, matriculated`, `grade 10-12, not matriculated`, or `post-secondary` |
| `sex` | Biological sex (all `F` — female) |

```{r c01}
participants_inst <- read_csv("../../data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv")
glimpse(participants_inst)
```

---

## `02_visit_clinical_measurements_UKZN_workshop_2023.csv`

**What a row represents:** one participant's clinical measurements at one study
visit.

**Columns:**

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID. Links to `00_sample_ids_UKZN_workshop_2023.csv` |
| `time_point` | Study visit: `baseline`, `week_1`, or `week_7` |
| `arm` | Study group: `treatment` or `placebo` |
| `nugent_score` | A score (0–10) of vaginal microbiome health — higher = more disturbed |
| `crp_blood` | C-reactive protein level in blood (mg/L) — a marker of inflammation |
| `ph` | Vaginal pH — normally acidic (~4.5), higher pH can indicate infection |

```{r c02}
clinical <- read_csv("../../data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv")
glimpse(clinical)
```

---

## `03_elisa_cytokines_UKZN_workshop_2023.csv`

**What a row represents:** one cytokine measurement for one sample (long format).
ELISA is a lab technique that measures protein concentrations using antibodies.

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_UKZN_workshop_2023.csv` |
| `cytokine` | Cytokine name: `IL-1a`, `IL-1b`, `IL-6`, `IL-8`, `IL-10`, `TNFa`, `IFN-Y`, `IP-10`, `MIG`, `MIP-3a` |
| `conc` | Concentration in pg/mL |
| `limits` | Measurement quality: `within limits` or `out of range` |

```{r c03}
elisa <- read_csv("../../data/instructional_dataset/03_elisa_cytokines_UKZN_workshop_2023.csv")
glimpse(elisa)
```

---

## `04_flow_cytometry_UKZN_workshop_2023.csv`

**What a row represents:** one sample's immune cell counts from flow cytometry.

**Columns:**

| Column | Meaning |
|--------|---------|
| `sample_id` | Unique sample identifier. Links to `00_sample_ids_UKZN_workshop_2023.csv` |
| `live_cd19_negative` | Number of live cells that are not B-cells |
| `cd45_negative` | Cells that are not immune cells (no CD45 marker) |
| `cd45_positive` | Immune cells (express CD45) |
| `neutrophils` | Neutrophil count |
| `non_neutrophils` | Immune cells that aren't neutrophils |
| `cd3_negative` | Non-T-cell immune cells |
| `cd3_positive` | T-cells |
| `cd4_t_cells` | Helper T-cells |
| `cd8_t_cells` | Killer T-cells |

```{r c04}
flow_inst <- read_csv("../../data/instructional_dataset/04_flow_cytometry_UKZN_workshop_2023.csv")
glimpse(flow_inst)
```

---

# Dataset D — Sorghum Timecourse (`data/modeling/01-sorghum_timecourse.csv`)

A plant biology dataset tracking **gene expression** in sorghum (a grain crop)
across a growing season. Two genotypes (BT642 and RT430) were grown under
different conditions (Control, Preflowering, Postflowering), and leaf samples
were collected weekly from Day 14 to Day 98.

The first nine columns are **gene expression values** — each is the expression
level of one specific sorghum gene in that sample.

## `01-sorghum_timecourse.csv`

**What a row represents:** one leaf sample collected at one time point from one
plant (or pool of plants), with expression measurements for nine genes.

**Columns:**

| Column | Meaning |
|--------|---------|
| `Sobic.001G062300.v3.1` through `Sobic.003G421700.v3.1` | Expression level of each sorghum gene (9 genes total). These are Sorghum bicolor gene IDs from genome version 3.1 |
| `Barcode` | Sequencing barcode used to identify this sample in the lab |
| `libraryName` | Sequencing library name (e.g. `BOCAX`) |
| `Block` | Field block number (e.g. `Block 05`) |
| `Week` | Week of the study (3–14) |
| `Replicate` | Biological replicate number (1–3) |
| `Genotype` | Sorghum variety: `BT642` or `RT430` |
| `Condition` | Growth condition: `Control`, `Preflowering`, or `Postflowering` |
| `Sample.type` | Type of tissue sampled (all `Leaves`) |
| `Day` | Day of the study (Day14 through Day98) |
| `Row` | Field row position |
| `isFloweringWeek` | Whether this sample was collected during the week the plant flowered (`TRUE`/`FALSE`) |
| `Date.Harvested` | Calendar date of harvest |
| `Collection.Time` | Time of day collected |
| `No.plants.pooled` | Number of plants pooled into this sample |
| `FloweringJulianDate` | Julian date (day of year) when the plant flowered |
| `FloweringDate` | Calendar date when the plant flowered |
| `FloweringWeek` | Week number when flowering occurred |
| `FloweringDay` | Day of study when flowering occurred |
| `Unreplicated` | Whether this sample had no replication (`TRUE`/`FALSE`) |
| `CompleteSampleTreatment` | Composite label combining Genotype, Week, and Condition (e.g. `RT430:03:Preflowering`) |
| `Timepoint` | Numeric week (same as `Week`) |
| `Group` | Composite label combining Genotype and Condition (e.g. `RT430:Preflowering`) |

```{r d01}
sorghum <- read_csv("../../data/modeling/01-sorghum_timecourse.csv")
glimpse(sorghum)
```